Validation of Visual Statistical Inference, Applied to Linear Models

### 2012-07-01 paper_jasa.tex
comment on choices
uncontaminated p-value plot added
worked on last two sections of paper - some questions to be answered

\begin{figure}[htp]
%\begin{figurehere}
   \centering
       \includegraphics[width=0.95\textwidth]{lineup_category.pdf}
       \caption{Lineup plot ($m=20$) using side-by-side boxplots for testing $H_0: \beta_k=0$. One of these plots is the plot of the actual data, and the remaining are null plots, produced by simulating data from a null model that assumes $H_0$ is true. Which plot is the most different from the others, in the sense that there is the the largest shift or location difference between the boxplots? (The position of the actual data plot is provided in Section \ref{sec:category}.)}
       \label{fig:test_category}
\end{figure}
%\end{figurehere}

@@

Using the inverse {\it logit} link function, $g^{-1}(.)$, from  Equation \ref{eqn:mixed} we obtain the the estimate of subject and lineup specific probability of successful evaluation as 
\begin{equation} \label{eqn:mixed_power}
\hat p_{\ell i} =  g^{-1}(W_{\ell i} \hat {\delta} +  Z_{\ell i}  \hat {\tau}_{\ell i}).
\end{equation}

@@

\begin{equation}\label{power_exp} 
   Power(\beta)=Pr(p_{D} < p_0)  \quad \text{for}  \quad \beta \ne 0
\end{equation}
Figure \ref{fig:power_expected} shows the power of the conventional test in comparison to the expected power of the visual test obtained from Equation \ref{power_exp}. 
Figure \ref{fig:power_expected} shows the power of the conventional test in comparison to the expected power of the visual test obtained from Equation \ref{power_exp}. Notice that the expected power of the visual test is %almost as good as 
close to the power of the conventional test.

@@

Three experiments were conducted to evaluate the effectiveness of the lineup protocol relative to the equivalent test statistic used in the regression scenario. There are three experiments. The first two are classical settings where we would not expect the lineup protocol to do better than the traditional test, but we hope it to perform favorably. The third experiment is a scenario where assumptions required for the traditional test are violated, and we would expect the lineup protocol to outperform the traditional test. (Data and lineups used in the experiments are available in the supplementary material.)

@@

\begin{figure*}[hbtp]
   \centering
       \scalebox{0.6}{\includegraphics{p_val_percent_correct.pdf}}
       \caption{Proportion of correct responses decreases rapidly with the increase of p-value. After p-value exceeds 0.15 it is very unlikely to identify the actual plot. The theoretical justification of this is shown in figure \ref{fig:pval_power}. }
       \label{fig:pval_pcorrect}
\end{figure*}

@@

\begin{figure*}[hbtp]
   \centering
       \scalebox{0.6}{\includegraphics{p_val_plot_signal.pdf}}
       \caption{Conventional test $p$-value ($p_D$) vs plot signal strength ($\hat p_D$). The solid line is a 45 degree line which represents the equity shown in Equation \ref{eqn:power_estimate}. p-values are shown on square root scale scale.}
    %   \hh{Mahbub, another change: could you change the y label to 'estimate of visual p-value $\hat{p_D}$ ?}
       \label{fig:pval_plot_signal}
\end{figure*}

@@

Another difference between conventional tests and visual inference is that the observer's choice of plot may be influenced by the $m-1$ null plots in the lineup. We have already seen that observers tend to pick the plot with the lowest $p$-value. If one or more null plots represent data having -- by chance -- a smaller $p$-value than the actual data plot, there's a strong possibility that an observer will pick this null plot instead of the actual data plot.

Figure \ref{fig:pval_difference} investigates this for the data collected in experiments 1 and 2. The horizontal axis shows the difference between the smalles $p$-value of the null plots, and that of the actual data plot. Large values indicate that the actual data plot has a much smaller $p$-value than any of the null plots. The vertical axis shows percent subjects choosing the actual data plot. \green{Is this really pct correct, or 1-pct correct? When there is a big diff in p-value almost no-one gets it right! Maybe it is just a scale issue?} We see that when there is a big difference between the $p$-values observers choose the actual data plot more often.

@@

This problem is further investigated in the paper \cite{waldo}. Here numerical measures of the similarity between plots are proposed to provide quality metrics for lineups.

@@

\subsection{Alternative - Type III error}~\label{sec:TypeIII}

\hh{From figure \ref{fig:power_loess_effect} we see that identification of the data plot is high independently of the effect size for experiment 3. This might be an indication that participants were not considering slope as distinguishing feature for the data, but the apparent clustering, i.e. the null hypothesis is being rejected, but not for the reason of the slope -- which the experimental setup was asking for --  leading to a type III error of correctly rejecting the null hypothesis for the wrong reason', as defined by \citet{mosteller:48}.}

@@

\blue{We also see this feature in figure \ref{fig:choice_reason}}.
\hh{Mahbub, it's not directly fig \ref{fig:choice_reason} - it's actually the lineup corresponding to the large outlier at (0.2,  0.6) in the left panel of fig \ref{fig:choice_reason}.}

@@

>>>>

\begin{figure}[htp]
%\begin{figurehere}
   \centering
       \includegraphics[width=0.95\textwidth]{lineup_category.pdf}
       \caption{Lineup plot ($m=20$) using side-by-side boxplots for testing $H_0: \beta_k=0$. One of these plots is the plot of the actual data, and the remaining are null plots, produced by simulating data from a null model that assumes $H_0$ is true. Which plot is the most different from the others, in the sense that there is the largest shift or location difference between the boxplots? (The position of the actual data plot is provided in Section \ref{sec:category}.)}
       \label{fig:test_category}
\end{figure}
%\end{figurehere}

@@

Using the inverse {\it logit} link function, $g^{-1}(.)$, from  Equation \ref{eqn:mixed} we obtain the estimate of subject and lineup specific probability of successful evaluation as 
\begin{equation} \label{eqn:mixed_power}
\hat p_{\ell i} =  g^{-1}(W_{\ell i} \hat {\delta} +  Z_{\ell i}  \hat {\tau}_{\ell i}).
\end{equation}

@@

\begin{equation}\label{power_exp} 
   Power(\beta)=Pr(p_{D} < p_0)  \quad \text{for}  \quad \beta \ne 0
\end{equation}
Figure \ref{fig:power_expected} shows the power of the conventional test in comparison to the expected power of the visual test obtained from Equation \ref{power_exp}. Notice that the expected power of the visual test is %almost as good as 
close to the power of the conventional test.

@@

Three experiments were conducted to evaluate the effectiveness of the lineup protocol relative to the equivalent test statistic used in the regression scenario. The first two are classical settings where we would not expect the lineup protocol to do better than the traditional test, but we hope it to perform favorably. The third experiment is a scenario where assumptions required for the traditional test are violated, and we would expect the lineup protocol to outperform the traditional test. (Data and lineups used in the experiments are available in the supplementary material.)

@@

\begin{figure*}[hbtp]
   \centering
       \scalebox{0.6}{\includegraphics{p_val_prop_correct.pdf}}
       \caption{Proportion of correct responses decreases rapidly with the increase of p-value. After p-value exceeds 0.15 it is very unlikely to identify the actual plot. The theoretical justification of this is shown in figure \ref{fig:pval_power}. }
       \label{fig:pval_pcorrect}
\end{figure*}

@@

\begin{figure*}[hbtp]
   \centering
       \scalebox{0.5}{\includegraphics{p_val_plot_signal.pdf}}
       \caption{Conventional test $p$-value ($p_D$) vs plot signal strength ($\hat p_D$). The solid line is a 45 degree line which represents the equity shown in Equation \ref{eqn:power_estimate}. p-values are shown on square root scale scale.}
    %   \hh{Mahbub, another change: could you change the y label to 'estimate of visual p-value $\hat{p_D}$ ?}
       \label{fig:pval_plot_signal}
\end{figure*}

@@

Visual inference is like other types of computationally intensive test procedures, such as permutation tests, where the statistics from the data are compared with those from null data. But unlike these approaches visual inference is constrained to make the comparison with just a few draws $(m-1)$ from the null distribution. How this small set of null plots influences the observers choice is important for understanding the reliability of visual inference. If the actual data plot is very different from the null plots, then the null plots should not have much influence on the choice. Measuring the difference between the plots is almost impossible! But in the controlled setting here we can again use $p$-value as a proxy for similarity and difference.

In the previous section we noticed that observers tend to pick the plot with the lowest $p$-value. If a null plot has data that, by chance because the sample is a random draw from the null, has a smaller $p$-value than the actual data, we would expect that there is a strong possibility that an observer will pick this null plot instead of the actual data plot. Figure \ref{fig:pval_difference} investigates this for the data collected in experiments 1 and 2. The plots display the proportion of observers who choose the actual data plot against the $p$-value of the actual data minus the lowest seen in the null data. A small negative value of the difference means that the $p$-value of the actual data is smaller than that of the null data. If there is a big negative difference we'd expect to see We would expect to see that the observers would more often make the correct choice of plot. This can be seen. When there is a large positive difference If however, there is little difference between the $p$-values then we'd expect to see more confusion in the choices.

\green{Something is definitely not right! The code looks like it does actual data p-value - min(null pvalue). But how can you get a difference of 0.8 from this??}

 \green{Is this really pct correct, or 1-pct correct? When there is a big diff in p-value almost no-one gets it right! Maybe it is just a scale issue?} We see that when there is a big difference between the $p$-values observers choose the actual data plot more often.

@@

In practice, $p$-value is not going to be a valid way to compare plots. Rather metrics that can measure how graphical elements from one plot to another are perceived similarly are needed. This is investigated in the paper \cite{waldo}. Here numerical measures of the similarity between plots are proposed to provide quality metrics for lineups.

@@

\subsection{Type III error}~\label{sec:TypeIII}

A little known error amongst statisticians is what was coined as Type III in \citet{mosteller:48}. This occurs when the null hypothesis is rejected for thw rong reason. Experiment 3 is prone to this type of error. Observers were asked to identify the plot that had the largest slope, either positive or negative. But the actual data plot had a cluster of points, the contamination that forced the conventional test to fail to see the association. For the human eye this cluster of points is probably as visible as the association between the remaining points, leading the observer to detect the actual data plot by looking for the cluster instead of the slope. This would be considered to be a Type III error because it is not what was requested, but it still gives the correct answer. This is not a problem with the lineup protocol, because we want observers to react to anything that sets the the actual data plot apart. Here, in the controlled but simple setting, it is of interest to see what observers cued to in the actual data plot.

Figure \ref{fig:power_loess_effect} indicates that Type III error might be occurring in experiment 3: correct identification of the actual data plot is not positively associated with effect size. Teasing this out of the results is possible by looking at the reasons observers gave for their choices. Observers were provided with four possible choices

\begin{enumerate} \itemsep 0in
\item Most different plot
\item Visible trend
\item Clustering visible
\item Other
\end{enumerate}

\noindent It was possible to choose more than one reason. The question begged for ``visible trend'' (2) to be the reasons that observers gave for correctly choosing the actual data plot. Either reasons 1 or 3 would be indicative of Type III error. It is interesting to calculate the proportion of observers who choose the correct plot, but use reasons 1 or 3, relative to reason 2. Figure \ref{fig:choice_reason} examines this. For observers who chose the actual data plot, the proportions for the three reasons are shown. It can be seen that the most common reason was ``most different plot'' (1), followed by clustering visible (3), with ``visible trend'' coming in third. \green{What about other??}

%This might be an indication that participants were not considering slope as distinguishing feature for the data, but the apparent clustering, i.e. the null hypothesis is being rejected, but not for the reason of the slope -- which the experimental setup was asking for --  leading to a type III error of correctly rejecting the null hypothesis for the wrong reason', as defined by \citet{mosteller:48}

@@

%\blue{We also see this feature in figure \ref{fig:choice_reason}}.
%\hh{Mahbub, it's not directly fig \ref{fig:choice_reason} - it's actually the lineup corresponding to the large outlier at (0.2,  0.6) in the left panel of fig \ref{fig:choice_reason}.}

\green{I find this plot really confusing! Can we just look at whether they selected choice ``clustering visible'' or not, if they were correct?}

@@

In practice there is no correct answer. In these experiments we controlled the effets that observers saw, so there was a correct choice in each case. In the real setting, there is not correct choice, just whether the actual data plot stands apart from the rest, so that the observer can find it. And observers would provide open-ended answers for their choice of plot.

What are your suggestions for future experiments, and next steps of the research?
